Performance for the masses: Experiments with A Web Based Architecture to Harness Volunteer Resources for Low Cost Distributed Evolutionary Computation

### 2016-02-03 GECCO/volunteer-clouds-gecco16.tex
More sections

On the other hand, we are also interested in measuring the performance
of volunteer computing systems, an area in which there have been
relatively few efforts. There were some initial attempts to avoid the differences in performance
that could be obtained from volunteers  by making
the algorithm adaptive to the kind of resources allotted to it
\cite{milani2004online}. This is actually not such a big problem in
algorithms such as the EA that can easily be
parallelized via population splitting or by farming out evaluations to all
available nodes. Lately, several approaches have focused on the
fault-tolerance of volunteer algorithms
\cite{gonzalez2010characterizing} which it can, of course, be studied in
the more general context of distributed computing 
\cite{nogueras2015studying} or be included in a more general study of the
performance of the EA itself
\cite{DBLP:journals/gpem/LaredoBGVAGF14}.

On the other hand, measuring the performance of the resulting metacomputer
involves understanding its dynamics. Initial
work was done for peer to peer systems by Stutzbach et
al. \cite{stutzbach2006understanding} and extended to volunteer
computing by Laredo et al. \cite{churn08,laredo2008rcp}. 
%But the raw material of           
%volunteer computing, % (Paloma) Whith regard to?  Maybe it was colloquial/analogy
But some of the essential metrics in volunteer computing like the
number of users or the time spent by every one in the
computation in browser-based volunteer computing experiments, have
only been studied in a limited way in 
\cite{DBLP:journals/gpem/LaredoBGVAGF14} on the basis of a single
run. Studies using volunteer computing platforms such as SETI@home
\cite{javadi2009mining} found out that the Weibull, log-normal, and
Gamma distribution 
modeled quite well the availability of resources in several clusters
of that framework; the shape of those distributions is a skewed bell
with more resources in the {\em low} areas than in the high areas:
there are many users that give a small amount of cycles, while there
are just a few that give many cycles. This is in concordance with the
results obtained in \cite{agajaj}. Although SETI@home works
at a larger scale in both quantity of volunteers and the scope of projects, 
our initial tests indicate that
this kind of distribution might be the most appropriate to describe
user participation in several different metrics. However, in this
paper we will perform different experiments in different setups and
will try to come up with a more precise model. Next we describe the
framework itself.

As far as we know, this paper presents one of the few experimental
results using a volunteer computing system which uses  % (Paloma) Change to "which" in order to not repeating "that" all the time
very diverse computational resources; these will be described next.

@@

We set up different experiments by deploying the server in free cloud
services, changing it from one experiment to the next if needed. This server runs a web service
which stores chromosomes sent from clients and serves requests to
them; it is 
an open source system that has been described extensively in
\cite{2016arXiv160101607Manom}. The client runs an evolutionary
algorithm written in JavaScript, that after a number of generations
sends back the best individual to the server, requesting at the same
time a random individual of those stored in the server. The experiment
starts by setting up a website that is sent by the server and announcing it on several social
networks so that people contribute to them by going to the page and
staying there for a while.

\subsection{Initial tests}

Initial tests with an evolutionary algorithm solving the 40-Trap
problem were set up using the free PaaS (Platform as a
Service) OpenShift; they are described in
Table \ref{tab:summary:os}. In general, every tests took several
days and during it the problem was solved more than 50 times with
several hundreds different IPs participating in it; the
evolutionary algorithm stopped when the solution was found. 
%
\begin{table*}
\caption{Summary of time per run, number of IPs and number of PUTs per IP in the initial runs. \label{tab:summary:os}}
\begin{center}
\begin{tabular}{l|ccccccccc}
\hline
Date & Experiments &\#IPs &\#IPs &  \#IPs & Time (s) &  \#PUTs & $<$ 69s & $<$ 3.46s & Inter-experiment\\
     & \# & Different & Median & Max & Median &  Median & & & correlation \\ 
\hline
4/4 & 57 & 191 & 5 & 16 & 2040 & 18 & 14.29\% & 5.36\% & 0.0082 \\
4/24 & 231 & 559 & 5 & 29 & 732 & 11 & 47.39\% & 3.91\% & 0.0934\\
7/31 & 97 & 179 & 5 & 14 & 260 & 23 & 27.08\% & 1.04\%  & 0.1741\\
\hline
\end{tabular}
\end{center}
\end{table*}
%

The last column shows the percentage of experiments in which running the
experiment in the browser takes {\em less} time than in the
desktop. Although volunteer computing systems do not exclude desktop
computing and, in fact, the server can handle any client, not only
those running JavaScript on the browser, we include it here for the
sake of comparison and to show that, being a free resource, it can
anyway make a valid and valuable contribution to any distributed or
desktop computing experiment. An interesting result is also that the
correlation between users of an experiment and the next is
non-existent, that is, computing nodes are not {\em
  staying} after the solution has been found. This is an issue that we
will address in this paper.

@@

At any rate, this also shows that a convenient way of increasing the
computing power would be increasing this minimum amount per user. This
issue led us to create the next version of the framework, which is
presented next. % (Paloma) This sentences are sort of alone here... Could we extend a bit the first "this" to connect them to the previous paragraph ?

@@

However, it is interesting to note why this is so, and the first hint
is the inter-experiment correlation between the number of IPs in
successive experiments. While before it was an unremarkable 10\%, it is
now more than 43\%, making the number of IPs in contiguous experiments
highly correlated. The main reason for this is the setup in {\sf
  NodIO-W$^2$}, which restarts an island after a solution has been
found and also that it keeps running even if the tab is not on the
foreground. This means that volunteers can leave the experiment
running for as long as they want and they will be contributing 
experiment after experiment, unlike the previous version, where they
stopped contributing after one solution was found. This is also
reflected in the number of PUTs, which is the number of generations
divided by 100; more than 50\% of the volunteers run the experiment for more than
4000 generations. All things considered, this means than there are
many more computers {\em simultaneously} running, leading to this
almost 5-fold increase in running time.

>>>>

On the other hand, we are also interested in measuring the performance
of volunteer computing systems, an area in which there have been
relatively few efforts. There were some initial attempts to avoid the differences in performance
that could be obtained from volunteers  by making
the algorithm adaptive to the kind of resources allotted to it
\cite{milani2004online}. This is actually not such a big problem in
algorithms such as the EA that can easily be
parallelized via population splitting or by farming out evaluations to all
available nodes. Lately, several approaches have focused on the
fault-tolerance of volunteer algorithms
\cite{gonzalez2010characterizing} which it can, of course, be studied in
the more general context of distributed computing 
\cite{nogueras2015studying} or be included in a broader study of the
performance of the EA itself
\cite{DBLP:journals/gpem/LaredoBGVAGF14}.

On the other hand, measuring the performance of the resulting metacomputer
involves understanding its dynamics. Initial
work was done for peer to peer systems by Stutzbach et
al. \cite{stutzbach2006understanding} and extended to volunteer
computing by Laredo et al. \cite{churn08,laredo2008rcp}. 
%But the raw material of           
%volunteer computing, % (Paloma) Whith regard to?  Maybe it was colloquial/analogy
But some of the essential metrics in volunteer computing like the
number of users or the time spent by every one in the
computation in browser-based volunteer computing experiments, have
only been studied in a limited way in 
\cite{DBLP:journals/gpem/LaredoBGVAGF14} on the basis of a single
run. Studies using volunteer computing platforms such as SETI@home
\cite{javadi2009mining} found out that the Weibull, log-normal, and
Gamma distributions 
modeled quite well the availability of resources in several clusters
of that framework; the shape of those distributions is a skewed bell
with more resources in the {\em low} areas than in the high areas:
there are many users that give a small amount of cycles, while there
are just a few that give many cycles. This is in concordance with the
results obtained in \cite{agajaj}. Although SETI@home works
at a larger scale in both quantity of volunteers and the scope of projects, 
our initial tests indicate that
this kind of distribution might be the most appropriate to describe
user participation in several different metrics. However, in this
paper we will perform different experiments in different setups and
will try to come up with a more precise model. 
%Mario: erase: Next we describe the framework itself.
As far as we know, this paper presents one of the few experimental
results using a volunteer computing system which uses  % (Paloma) Change to "which" in order to not repeating "that" all the time
very diverse computational resources; these will be described next.

@@

We set up different experiments by deploying the server in free cloud
services, changing it from one experiment to the next if needed. This server runs a web service
which stores chromosomes sent from clients and serves requests to
them; it is 
an open source system that has been described extensively in
\cite{2016arXiv160101607Manom}. The client runs an evolutionary
algorithm written in JavaScript, that after a number of generations
sends the best individual back to the server, requesting at the same
time a random individual of those stored in the server. The experiment
starts by setting up a website that is sent by the server and announcing it on several social
networks so that people contribute to them by going to the page and
staying there for a while.

\subsection{Initial tests with a simple Ajax client}
%% Above we mention three experiments:  Simple Ajax client, Web Workers, Harder Problem
%% Im renaming subsection to  reference the first experiment

Initial tests with an evolutionary algorithm solving the 40-Trap
problem were set up using the free PaaS OpenShift; they are described in
Table \ref{tab:summary:os}. In general, every tests took several
days and during it the problem was solved more than 50 times with
several hundreds different IPs participating in it; each
evolutionary algorithm stopped when the solution was found. 
%
\begin{table*}
\caption{Summary of time per run, number of IPs and number of PUTs per IP in the initial runs. \label{tab:summary:os}}
\begin{center}
\begin{tabular}{l|ccccccccc}
\hline
Date & Experiments &\#IPs &\#IPs &  \#IPs & Time (s) &  \#PUTs & $<$ 69s & $<$ 3.46s & Inter-experiment\\
     & \# & Different & Median & Max & Median &  Median & & & correlation \\ 
\hline
4/4 & 57 & 191 & 5 & 16 & 2040 & 18 & 14.29\% & 5.36\% & 0.0082 \\
4/24 & 231 & 559 & 5 & 29 & 732 & 11 & 47.39\% & 3.91\% & 0.0934\\
7/31 & 97 & 179 & 5 & 14 & 260 & 23 & 27.08\% & 1.04\%  & 0.1741\\
\hline
\end{tabular}
\end{center}
\end{table*}
%

The last column shows the percentage of experiments in which running the
experiment in the browser takes {\em less} time than in the
desktop. Although volunteer computing systems do not exclude desktop
computing and, in fact, the server can handle any client, not only
those running JavaScript on the browser, we include it here for the
sake of comparison and to show that, being a free resource, it can
anyway make a valid and valuable contribution to any distributed or
desktop computing experiment. An interesting result is also that the
correlation between users of an experiment and the next is
non-existent, that is, computing nodes are not {\em
  staying} after the solution has been found. This is the issue that we
address in the next experiments. 
%Mario: changed: will address in this paper.

@@

At any rate, this also shows that a convenient way of increasing the
%Mario: As I understend is: 
computing power would be increasing this minimum amount of experiments per user. This
issue led us to create the next version of the framework, which is
presented next. % (Paloma) This sentences are sort of alone here... Could we extend a bit the first "this" to connect them to the previous paragraph ?

@@

However, it is interesting to note why this is so, and the first hint
is the inter-experiment correlation between the number of IPs in
successive experiments. While before it was an unremarkable 10\%, it is
now more than 43\%, making the number of IPs in contiguous experiments
highly correlated. The main reason for this is the setup in {\sf
  NodIO-W$^2$}, which restarts an island after a solution has been
found and also that it keeps running even if the tab is not on the
foreground. This means that volunteers can leave the experiment
running for as long as they want and they will be contributing 
experiment after experiment, unlike the previous version, where they
stopped contributing after one solution was found. This is also
reflected in the number of PUTs, which is the number of generations
divided by 100; more than 50\% of the volunteers run the experiment for more than
4000 generations. All things considered, this means than there are
%Mario: many more more threads {\em simultaneously} running?
many more computers {\em simultaneously} running, leading to this
almost 5-fold increase in running time.
